Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-09-29 ms/main.tex
Add analysis for T_A > T_B before result 3
moved tables to end; revised new section by DC; started new discussion par to summarize corollaries

%%% Table: payoff matrix
\bigskip
\begin{table}[h]
\centering
\caption{\textbf{Payoff matrix for prisoner's dilemma.}
The fitness of phenotype $\phi_1$ when interacting with phenotype $\phi_2$. $A$ is a cooperative phenotype, $B$ is a defector phenotype, $b$ is the benefit gained by an individual interacting with a cooperator, and $c$ is the cost of cooperation. $b>c>0$.
}
\begin{tabular}{lll}
\toprule
           & $\phi_2=A$ & $\phi_2=B$ \\ \cmidrule(r){1-3}
$\phi_1=A$ & $1+b-c$ & $1-c$ \\
$\phi_1=B$ & $1+b$   & $1$
\\ \bottomrule
\end{tabular}
\label{table:prisoner_payoff}
\end{table}
\bigskip
%%%

@@

%%% Table: interactions
\bigskip
\begin{table}[]
\caption{\textbf{Interaction frequency, fitness, and transmission probabilities.}}
\begin{tabular}{@{}llllll@{}}
\toprule
\multirow{2}{*}{Phenotype $\phi_1$} &
  \multirow{2}{*}{Phenotype $\phi_2$} &
  \multirow{2}{*}{Frequency} &
  \multirow{2}{*}{Fitness of $\phi_1$} &
  \multicolumn{2}{l}{$P(\phi_1=A)$ via horizontal transmission:} \\ \cmidrule(l){5-6} 
    &     &                      &         & from partner, $\alpha$ & from population, $(1-\alpha)$ \\ \cmidrule(r){1-6}
$A$ & $A$ & $\hat{p}^2$          & $1+b-c$ & 1                      & $\hat{p}+(1-\hat{p})(1-T_B)$  \\
$A$ & $B$ & $\hat{p}(1-\hat{p})$ & $1-c$   & $1-T_B$                & $\hat{p}+(1-\hat{p})(1-T_B)$  \\
$B$ & $A$ & $\hat{p}(1-\hat{p})$ & $1+b$   & $T_A$                  & $\hat{p} T_A$                 \\
$B$ & $B$ & $(1-\hat{p})^2$      & $1$     & $0$                    & $\hat{p} T_A$                 \\ \bottomrule
\end{tabular}
\label{table:interactions}
\end{table}
\bigskip
%%%

@@

These conditions are illustrated in \autoref{fig:results2}.

Cooperation and defection can coexist stably at frequencies $\tilde{p}^*$ and $1-\tilde{p}^*$ (\autoref{eq:vert_hori_equilibrium}) if there is horizontal transmission bias for defection, $T_A<T_B$, and the cost of cooperation is intermediate, $\gamma_2 < c < \gamma_1$; see yellow area in \autoref{fig:results2}b.
When unstable coexistence occurs, phenotype $A$ will fix if its initial frequency is $p>\tilde{p}^*$, and phenotype $B$ will fix if its initial frequency is $1-p>1-\tilde{p}^*$; this occurs when there is horizontal transmission bias for cooperation, $T_A>T_B$, and the cost is intermediate, $\gamma_1 < c < \gamma_2$.
\autoref{fig:coexistence_without_oblique} shows the mapping $\tilde p \to \tilde p'$; see blue areas in \autoref{fig:results2}a and \autoref{fig:results2}b.
\\

\begin{corollary}[Symmetric horizontal transmission] \label{sec:sym_hor_trans}
If $T=T_A=T_B$, then cooperation will take over the population if
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \alpha \frac{T}{1-T} \;.
\end{equation}
\end{corollary}
Note that the right-hand side equals $\gamma_1$ when $T=T_A=T_B$.
This condition is obtained by setting $T=T_A=T_B$ in inequality \ref{eq:vert_hori_global_condition} and can be interpreted as a version of Hamilton's rule  ($c<b\cdot r$, inequality~\ref{eq:hamilton_rule}) or as a version of inequality~\ref{eq:fletcher2009}, where $\alpha T/(1-T)$ can be regarded as the \emph{effective relatedness} or \emph{effective assortment}, respectively.
\autoref{fig:results}a illustrates this condition.

@@

\autoref{fig:results2}a illustrates these conditions, which are obtained by setting $\alpha=0$ in Eqs.\ \ref{eq:vert_hori_equilibrium} and \ref{eq:cost_boundaries}. If there is horizontal transmission bias for cooperation ($T_A>T_B$) then cooperation can evolve from any initial frequency if the cost is low enough, $c<\gamma_1$ (green area below solid line). If the cost is not low enough, cooperation can also evolve if its initial frequency is high enough and the cost is not too high, $c<\gamma_2$ (blue area between solid and dashed lines).

@@

\autoref{fig:results2}c demonstrates these conditions.
With horizontal transmission bias for cooperation ($T_A>T_B$), cooperation can fix from any initial frequency if $a_2<\alpha$ (green area in positive x-axis). 
Otherwise, cooperation can also fix if the horizontal bias is strong, the initial frequency is high, and social association is high, $a_1 < \alpha$ (blue area).
With horizontal bias favoring defection ($T_A<T_B$), cooperation can fix from any frequency if social association is high, $a_1<\alpha$ (green area with $T_A<T_B$), and can also increase when rare and reach stable coexistence with defection if social association is intermediate, $a_2<\alpha$ (yellow area).
Without horizontal bias ($T_A=T_B$) fixation of cooperation occurs if social association is high enough; $\frac{c}{b} \cdot \frac{1-T}{T} < \alpha$ (inequality~\ref{eq:equal_transmission}; in this case $a_1=a_2$).

@@

If $T=T_A=T_B$ then $\beta_1=0$ and $\beta_3=-\beta_2=\alpha b vT -cv(1-T)$, 
and $f(\hat{p})$ becomes a quadratic polynomial:
\begin{equation} \label{eq:equal_horizontal_transmission}
  f(\hat{p}) = \hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].
\end{equation}
Clearly the only two equilibria are the fixations  $\hat{p} =  0$ and $\hat{p} = 1$.
These equilibria are locally stable if $f'(\hat{p})<0$ near the equilibrium (see Appendix~\autoref{sec:appendixB}), where
\begin{equation}
f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
\end{equation}
with
\begin{equation} \label{eq:derivative_of_phattag-phat}
\begin{aligned}
	f'(0) &=	\alpha bvT - cv(1-T), \\
	f'(1) &=	-\alpha bvT + cv(1-T).
\end{aligned}
\end{equation}
Therefore with symmetric horizontal transmission ($T_A=T_B$), fixation of the cooperative phenotype ($\hat{p}=1$) occurs under the same condition as Corollary 1.1, namely Eq.\ \ref{eq:equal_transmission}.

@@

Now, assume $T_A<T_B$.
$\beta_3<0$ requires $\hat\gamma_1<c$,
and for $\beta_3<\beta_1$ we need $c\big[v(1-T_B) + (1-v)(T_A-T_B)\big] > bv\alpha T_B + (1+b)(T_A-T_B)$.
Note that the expression in the square brackets is positive if and only if $v > \hat v$.
Thus, for $\beta_3<\beta_1$ we need $v > \hat v$ and $\hat\gamma_2 < c$ or $v < \hat v$ and $c < \hat\gamma_2$,
and for $0<\beta_3<\beta_1$ we need $v > \hat v$ and $\hat\gamma_2 < c < \hat\gamma_1$, or $v < \hat v$ and $c < \min(\hat\gamma_1, \hat\gamma_2)$. For $\beta_1<\beta_3$ we need $v > \hat v$ and $c<\hat\gamma_2$ or $v < \hat v$ and $\hat\gamma_2<c$.
However, some of these conditions cannot be met, since $v < \hat v$ implies $c<1<\hat\gamma_2$.
A similar analysis can be made when $T_A>T_B$.

@@

In general, this condition cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, i.e., with $T=T_A=T_B$, these conditions reduce to a form of Hamilton's rule.
\\
\begin{corollary}[Necessary condition for fixation of cooperation]
  Fixation of cooperation is possible only if the vertical transmission rate is high enough,
  \begin{equation} \label{eq:fixation_of_cooperation_vertical_transmission_condition}
    \begin{aligned}
      v>\hat{v} = \frac{T_B - T_A}{1-T_A} \;.
    \end{aligned}
    \end{equation} 
\end{corollary}
%Note that $\sigma \frac{\partial \gamma_1}{\partial v} = \sigma (T_B-T_A)$ and $\sigma \frac{\partial \hat\gamma_2}{\partial v} = \sigma (T_A-T_B)(b \alpha T_B - (1+b)(1-T_A))$
Note that this condition does not guarantee fixation of cooperation. Rather, if this condition does not apply then cooperation cannot fix.
When horizontal transmission is biased for cooperation, $T_A>T_B$, cooperation can fix with any vertical transmission rate, i.e. $\hat{v}<0$.
In contrast, if the horizontal transmission is biased for defection, $T_A<T_B$,  cooperation can fix only if the vertical transmission rate is high enough: in this case oblique transmission can prevent fixation of cooperation, see \autoref{fig:result3}bd.
\\

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion
\section*{Discussion}
We studied the evolution of cooperation under non-vertical transmission using deterministic discrete-time evolutionary models with fitnesses in the form of payoffs from a prisoner's dilemma game. 
Under oblique and horizontal cultural transmission, horizontal transmission bias in favor of the cooperative phenotype was found to be necessary and sufficient for evolution of cooperation (Result~\autoref{result:obli_hori}).
Under a combination of vertical, oblique, and horizontal transmission, cooperation or defection can either fix or coexist at a stable polymorphism, depending on the relationship between the cost and benefit of cooperation, the horizontal bias, and the association between cooperation and transmission (Results~\autoref{result:vert_hori} and~\autoref{result:vert_obli_hori}).
Remarkably, stable coexistence between cooperation and defection can be maintained if horizontal transmission is biased for defection ($T_A<T_B$) and both the cost of cooperation and social association are intermediate (yellow areas in \autoref{fig:results2}).
% TODO figure for result 3
% TODO something about v-hat

We tested the applicability of our analytic results. Our derived analytic condition provided a good approximation to simulation results for a complex spatial stochastic model  (\autoref{fig:spatial}), demonstrating that association between cooperation and transmission can arise in a structured population simply because the choice of individuals for social interactions is very restricted.

@@

To understand the role of horizontal transmission, we first discuss the role of \emph{assortment}.
\citet{Eshel1982} showed that altruism can evolve when there is enough \emph{assortative meeting}, namely, a tendency for individuals to interact within their phenotypic group.
\citet{Fletcher2009assortment}  further argued that a general explanation for the evolution of altruism is given by \emph{assortment}: the correlation between individuals that carry an altruistic trait and the amount of altruistic behavior in their interaction group (see also \citet{Bijma2010assortment}).
They therefore suggested that to explain the evolution of altruism, we should seek mechanisms that generate  assortment, such as population structure, repeated interactions, and individual recognition.
Our results highlight another mechanism for generating assortment: an association between social interactions and horizontal transmission that creates a correlation between one's partner for interaction and partner for transmission.
This mechanism does not require population structure, repeated interactions, or individual recognition.
We show that high levels of such social association greatly increase the potential for evolution of cooperation (\autoref{fig:results2}).
With enough social association ($\alpha > \big(c(1-T_B) + (T_B-T_A)\big) / b T_A$), cooperation can increase in frequency when initially rare even when there is horizontal transmission bias against it ($T_A<T_B$, see Result~\autoref{result:vert_hori}).

@@

\section*{Figures}
%%% Figure: model illustration
\begin{figure}[thb]
  \centering
  \includegraphics[scale=1]{figure1.pdf}
  \caption{\textbf{Cultural horizontal transmission with assortment.} Transmission occurs between interacting partners with probability $\alpha$ (left) or between two random peers with probability $1-\alpha$, where $\alpha$ is the \emph{social association} parameter.
  }
  \label{fig:horizontal}
\end{figure}

%%% Figure: coexistence map - without oblique transmission
\begin{figure}[htb]
\centering
\includegraphics[scale = 0.75]{coexistence_without_oblique.pdf}
\caption{\textbf{Stable and unstable coexistence between cooperation and defection without oblique transmission.}
  The curves show the frequency of the cooperative phenotype $A$ among parents in the next generation, $\tilde{p}'$, vs.\ that in the current generation $\tilde{p}$ (\autoref{eq:nextgen_parents_vertical_only}).
  The dashed black line is $\tilde{p}'=\tilde{p}$.
  The curves and the dashed line intersect at the polymorphic equilibrium $\tilde{p}^*$ (black circle).
  When the curves are above the dashed line, $\tilde{p}'>\tilde{p}$, and $\tilde{p}$ increases.
  When the curves are below the dashed line, $\tilde{p}'<\tilde{p}$, and $\tilde{p}$ decreases.
  The orange curve, for which the polymorphic equilibrium is stable, is given by $T_A = 0.4$, $T_B = 0.9$, $b = 12$, $c=0.35$, and $\alpha = 0.45$, which give $\gamma_2<c<\gamma_1$ (\autoref{eq:cost_boundaries})
  The blue curve, for which the equilibrium is unstable, is given by $T_A = 0.5$, $T_B = 0.1$, $b = 1.3$, $c=0.904$, and $\alpha = 0.4$, which give $\gamma_1<c<\gamma_2$.
  In both cases there is no oblique transmission, $v=1$; see \autoref{fig:coexistence_with_oblique} for $v<1$.
  }
\label{fig:coexistence_without_oblique}
\end{figure}

@@

\begin{figure}[htb]
  \centering       
    \includegraphics[width=0.45\textwidth]{Result2_c_zero_alpha.eps}
	~
    \includegraphics[width=0.45\textwidth]{Result2_c_non_zero_alpha.eps}
    ~
    \includegraphics[width=0.45\textwidth]{Result2_alpha.eps}

    \caption{\textbf{Evolution of cooperation under vertical and horizontal cultural transmission.} 
The figure shows the global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable coexistence (blue), and stable coexistence of cooperation and defection (yellow).
In all cases the horizontal bias ($T_A-T_B$) is on the x-axis.
(\textbf{a-b})~The cost of cooperation $c$ is on the y-axis; the cost boundaries $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_boundaries}) are the solid and dashed lines, respectively. 
(\textbf{c})~social association $\alpha$ is on the y-axis; the social association boundaries $a_1$ and $a_2$ (\autoref{eq:boundries_assortative_meeting}) are the solid and dashed lines, respectively.
Here, $b=1.3$, $T_A=0.4$. (\textbf{a})~$\alpha = 0$.(\textbf{b})~$\alpha = 0.7$. (\textbf{c})~$c = 0.35$.
  	}
    \label{fig:results2}
\end{figure}
\pagebreak

@@

\begin{figure}[htb]
  \centering       
    \includegraphics[width=0.45\textwidth]{Result3_TA_TB.eps}~
    \includegraphics[width=0.45\textwidth]{Result3_TB_TA.eps}
    \includegraphics[width=0.45\textwidth]{Result3_alpha_Vs_v_TA_TB.eps}~
    \includegraphics[width=0.45\textwidth]{Result3_alpha_Vs_v_TB_TA.eps}
    \caption{\textbf{Evolution of cooperation under vertical, oblique, and horizontal cultural transmission.} 
    The figure shows the global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable coexistence (blue), and stable coexistence of cooperation and defection (yellow).
	In all cases the vertical transmission rate $v$ is on the x-axis.
	(\textbf{a-b}) The cost of cooperation $c$ is on the y-axis and the cost boundaries $\hat\gamma_1$ and $\hat\gamma_2$ (\autoref{eq:cost_boundaries_v}) are represented by the solid and dashed lines, respectively. 
    (\textbf{c-d}) The social association $\alpha$ is on the y-axis and the social association boundaries $\hat{a}_1$ and $\hat{a}_2$ (\autoref{eq:boundries_assortative_meeting_general_case}) are represented by the solid and dashed lines, respectively. 
    Horizontal transmission is biased in (\textbf{a,c}) for cooperation, $T_A>T_B$, and in (\textbf{b,d}) for defection, $T_A<T_B$.    
    Here, $T_A = 0.5$, and
    (\textbf{a}) $b=1.2$, , $T_B = 0.4$, $\alpha = 0.4$;
    (\textbf{b}) $b=2$, $T_B = 0.7$, $\alpha = 0.7$;
    (\textbf{c}) $b=1.2$, $T_B = 0.4$, $c=0.5$;
    (\textbf{d}) $b=2$, $T_B = 0.7$, $c=0.5$.
    }
    \label{fig:result3}
\end{figure}
\pagebreak

>>>>

@@

@@

These conditions are illustrated in \autoref{fig:result2}.

Cooperation and defection can coexist stably at frequencies $\tilde{p}^*$ and $1-\tilde{p}^*$ (\autoref{eq:vert_hori_equilibrium}) if there is horizontal transmission bias for defection, $T_A<T_B$, and the cost of cooperation is intermediate, $\gamma_2 < c < \gamma_1$; see yellow area in \autoref{fig:result2}b.
When unstable coexistence occurs, phenotype $A$ will fix if its initial frequency is $p>\tilde{p}^*$, and phenotype $B$ will fix if its initial frequency is $1-p>1-\tilde{p}^*$; this occurs when there is horizontal transmission bias for cooperation, $T_A>T_B$, and the cost is intermediate, $\gamma_1 < c < \gamma_2$.
\autoref{fig:coexistence_without_oblique} shows the mapping $\tilde p \to \tilde p'$; see blue areas in \autoref{fig:result2}a and \autoref{fig:result2}b.
\\

\begin{corollary}[Symmetric horizontal transmission] 
\label{sec:cor_equal_transmission}
If $T=T_A=T_B$, then cooperation will take over the population if
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \alpha \frac{T}{1-T} \;.
\end{equation}
\end{corollary}
Note that the right-hand side equals $\gamma_1$ when $T=T_A=T_B$.
This condition is obtained by setting $T=T_A=T_B$ in inequality \ref{eq:vert_hori_global_condition} and can be interpreted as a version of Hamilton's rule  ($c<b\cdot r$, inequality~\ref{eq:hamilton_rule}) or as a version of inequality~\ref{eq:fletcher2009}, where $\alpha T/(1-T)$ can be regarded as the \emph{effective relatedness} or \emph{effective assortment}, respectively.
\autoref{fig:results}a illustrates this condition.

@@

\autoref{fig:result2}a illustrates these conditions, which are obtained by setting $\alpha=0$ in Eqs.\ \ref{eq:vert_hori_equilibrium} and \ref{eq:cost_boundaries}. If there is horizontal transmission bias for cooperation ($T_A>T_B$) then cooperation can evolve from any initial frequency if the cost is low enough, $c<\gamma_1$ (green area below solid line). If the cost is not low enough, cooperation can also evolve if its initial frequency is high enough and the cost is not too high, $c<\gamma_2$ (blue area between solid and dashed lines).

@@

\autoref{fig:result2}c demonstrates these conditions.
With horizontal transmission bias for cooperation ($T_A>T_B$), cooperation can fix from any initial frequency if $a_2<\alpha$ (green area in positive x-axis). 
Otherwise, cooperation can also fix if the horizontal bias is strong, the initial frequency is high, and social association is high, $a_1 < \alpha$ (blue area).
With horizontal bias favoring defection ($T_A<T_B$), cooperation can fix from any frequency if social association is high, $a_1<\alpha$ (green area with $T_A<T_B$), and can also increase when rare and reach stable coexistence with defection if social association is intermediate, $a_2<\alpha$ (yellow area).
Without horizontal bias ($T_A=T_B$) fixation of cooperation occurs if social association is high enough; $\frac{c}{b} \cdot \frac{1-T}{T} < \alpha$ (inequality~\ref{eq:equal_transmission}; in this case $a_1=a_2$).

@@

If $T=T_A=T_B$ then $\beta_1=0$ and $\beta_3=-\beta_2=\alpha b vT -cv(1-T)$, 
and $f(\hat{p})$ becomes a quadratic polynomial:
\begin{equation} \label{eq:equal_horizontal_transmission}
  f(\hat{p}) = \hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].
\end{equation}
Clearly the only two equilibria are the fixations  $\hat{p} =  0$ and $\hat{p} = 1$.
These equilibria are locally stable if $f'(\hat{p})<0$ near the equilibrium (see Appendix~\autoref{sec:appendixB}), where
\begin{equation}
f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
\end{equation}
with
\begin{equation} \label{eq:derivative_of_phattag-phat}
\begin{aligned}
	f'(0) &=	\alpha bvT - cv(1-T), \\
	f'(1) &=	-\alpha bvT + cv(1-T).
\end{aligned}
\end{equation}
Therefore with symmetric horizontal transmission ($T_A=T_B$), fixation of the cooperative phenotype ($\hat{p}=1$) occurs under the same condition as Corollary \ref{sec:cor_equal_transmission}, namely inequality~\ref{eq:equal_transmission}.

@@

First, assume $T_A<T_B$.
$\beta_3<0$ requires $\hat\gamma_1<c$,
and for $\beta_3<\beta_1$ we need $c\big[v(1-T_B) + (1-v)(T_A-T_B)\big] > bv\alpha T_B + (1+b)(T_A-T_B)$.
Note that the expression in the square brackets is positive if and only if $v > \hat v$.
Thus, for $\beta_3<\beta_1$ we need $v > \hat v$ and $\hat\gamma_2 < c$ or $v < \hat v$ and $c < \hat\gamma_2$,
and for $0<\beta_3<\beta_1$ we need $v > \hat v$ and $\hat\gamma_2 < c < \hat\gamma_1$, or $v < \hat v$ and $c < \min(\hat\gamma_1, \hat\gamma_2)$. For $\beta_1<\beta_3$ we need $v > \hat v$ and $c<\hat\gamma_2$ or $v < \hat v$ and $\hat\gamma_2<c$.
However, some of these conditions cannot be met, since $v < \hat v$ implies $c<1<\hat\gamma_2$.

Second, when $T_A>T_B$
$\beta_3>0$ requires $\hat\gamma_1 > c$. 
For $\beta_1<\beta_3$ we need $c\big[v(1-T_B) + (1-v)(T_A-T_B)\big] < bv\alpha T_B + (1+b)(T_A-T_B)$.
Thus for $\beta_1<\beta_3$ we need $v > \hat v$ and $c < \hat\gamma_2 $ or $v < \hat v$ and $c > \hat\gamma_2$.
But $\hat{v}<0$ when $T_A > T_B$, and therefore we have $\beta_1<\beta_3$ if $c < \hat\gamma_2$. Similarly, we have $\beta_3<\beta_1$ if $c > \hat{\gamma_2}$.

@@

In general, this condition cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias $T_A-T_B$.
Without horizontal transmission bias, i.e., with $T=T_A=T_B$, these conditions reduce to a form of Hamilton's rule.
\\
\begin{corollary}[Necessary condition for fixation of cooperation]
\label{sec:cor_vhat}
  Fixation of cooperation is possible only if the vertical transmission rate is high enough,
  \begin{equation} \label{eq:fixation_of_cooperation_vertical_transmission_condition}
    \begin{aligned}
      v>\hat{v} = \frac{T_B - T_A}{1-T_A} \;.
    \end{aligned}
    \end{equation} 
\end{corollary}
%Note that $\sigma \frac{\partial \gamma_1}{\partial v} = \sigma (T_B-T_A)$ and $\sigma \frac{\partial \hat\gamma_2}{\partial v} = \sigma (T_A-T_B)(b \alpha T_B - (1+b)(1-T_A))$
Note that this condition does not guarantee fixation of cooperation. Rather, if this condition does not apply then cooperation cannot fix.
When horizontal transmission is biased for cooperation, $T_A>T_B$, cooperation can fix with any vertical transmission rate, i.e. $\hat{v}<0$.
In contrast, if the horizontal transmission is biased for defection, $T_A<T_B$,  cooperation can fix only if the vertical transmission rate is high enough: in this case oblique transmission can prevent fixation of cooperation, see \autoref{fig:result3}bd.
\\

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion
\section*{Discussion}
We studied the evolution of cooperation under non-vertical transmission using deterministic discrete-time evolutionary models with fitnesses in the form of payoffs from a prisoner's dilemma game. 
Under oblique and horizontal cultural transmission, horizontal transmission bias in favor of the cooperative phenotype was found to be necessary and sufficient for evolution of cooperation (Result~\autoref{result:obli_hori}).
Under a combination of vertical, oblique, and horizontal transmission, cooperation or defection can either fix or coexist at a stable polymorphism, depending on the relationship between the cost and benefit of cooperation, the horizontal bias, and the association between cooperation and transmission (Results~\autoref{result:vert_hori} and~\autoref{result:vert_obli_hori}).
Remarkably, stable coexistence between cooperation and defection can be maintained if horizontal transmission is biased for defection ($T_A<T_B$) and both the cost of cooperation and social association are intermediate (yellow areas in Figures~\ref{fig:result2}-\ref{fig:result3}).

We find that increasing social association ($\alpha$) increases the opportunity for evolution of cooperation (corollary X, \autoref{fig:result2}c and \autoref{fig:result3}cd). 
Similarly, a horizontal transmission bias for cooperation ($T_A-T_B$) will also facilitate its evolution (corollary X, \autoref{fig:result2}}.
The situation with vertical/oblique transmission ($v$) is more complex. 
When there is horizontal transmission bias for cooperation ($T_A>T_B$), increasing the rate of oblique transmission ($1-v$) will facilitate the evolution of cooperation (\autoref{fig:result3}ac). However, when the bias if for defection ($T_A<T_B$), high rates of vertical transmission ($v$) are advantageous for cooperation. 
Nevertheless, to reach fixation of cooperation in the population, there must be a sufficient rate of vertical transmission ($v>\hat v$, Corollary~\autoref{sec:cor_vhat}).
% TODO carry on summarizing corollaries here

Our analytic conditions derived for a deterministic well-mixed model provides a good approximation to results of simulations of a complex spatial stochastic model  (\autoref{fig:spatial}).
In this spatial model, individuals can only interact with and transmit to their neighbors on a grid.
This demonstrates that social association between cooperation and transmission can arise in a structured population in which interactions are local.

@@

To understand the role of horizontal transmission, we first discuss the role of \emph{assortment}.
\citet{Eshel1982} showed that altruism can evolve when there is enough \emph{assortative meeting}, namely, a tendency for individuals to interact within their phenotypic group.
\citet{Fletcher2009assortment}  further argued that a general explanation for the evolution of altruism is given by \emph{assortment}: the correlation between individuals that carry an altruistic trait and the amount of altruistic behavior in their interaction group (see also \citet{Bijma2010assortment}).
They therefore suggested that to explain the evolution of altruism, we should seek mechanisms that generate  assortment, such as population structure, repeated interactions, and individual recognition.
Our results highlight another mechanism for generating assortment: an association between social interactions and horizontal transmission that creates a correlation between one's partner for interaction and partner for transmission.
This mechanism does not require population structure, repeated interactions, or individual recognition.
We show that high levels of such social association greatly increase the potential for evolution of cooperation (\autoref{fig:result2}).
With enough social association ($\alpha > \big(c(1-T_B) + (T_B-T_A)\big) / b T_A$), cooperation can increase in frequency when initially rare even when there is horizontal transmission bias against it ($T_A<T_B$, see Result~\autoref{result:vert_hori}).

@@

\section*{Tables}

%%% Table: payoff matrix
\bigskip
\begin{table}[h]
\centering
\caption{\textbf{Payoff matrix for prisoner's dilemma.}}
\begin{tabular}{lll}
\toprule
           & $\phi_2=A$ & $\phi_2=B$ \\ \cmidrule(r){1-3}
$\phi_1=A$ & $1+b-c$ & $1-c$ \\
$\phi_1=B$ & $1+b$   & $1$
\\ \bottomrule
\end{tabular}
\label{table:prisoner_payoff}
\caption*{
The fitness of phenotype $\phi_1$ when interacting with phenotype $\phi_2$. $A$ is a cooperative phenotype, $B$ is a defector phenotype, $b$ is the benefit gained by an individual interacting with a cooperator, and $c$ is the cost of cooperation. $0<b<c$.
}
\end{table}
\bigskip
%%%

%%% Table: interactions
\bigskip
\begin{table}[h]
\caption{\textbf{Interaction frequency, fitness, and transmission probabilities.}}
\begin{tabular}{@{}llllll@{}}
\toprule
\multirow{2}{*}{Phenotype $\phi_1$} &
  \multirow{2}{*}{Phenotype $\phi_2$} &
  \multirow{2}{*}{Frequency} &
  \multirow{2}{*}{Fitness of $\phi_1$} &
  \multicolumn{2}{l}{$P(\phi_1=A)$ via horizontal transmission:} \\ \cmidrule(l){5-6} 
    &     &                      &         & from partner, $\alpha$ & from population, $(1-\alpha)$ \\ \cmidrule(r){1-6}
$A$ & $A$ & $\hat{p}^2$          & $1+b-c$ & 1                      & $\hat{p}+(1-\hat{p})(1-T_B)$  \\
$A$ & $B$ & $\hat{p}(1-\hat{p})$ & $1-c$   & $1-T_B$                & $\hat{p}+(1-\hat{p})(1-T_B)$  \\
$B$ & $A$ & $\hat{p}(1-\hat{p})$ & $1+b$   & $T_A$                  & $\hat{p} T_A$                 \\
$B$ & $B$ & $(1-\hat{p})^2$      & $1$     & $0$                    & $\hat{p} T_A$                 \\ \bottomrule
\end{tabular}
\label{table:interactions}
\end{table}
\bigskip
%%%

\section*{Figures}
%%% Figure: model illustration
\begin{figure}[h]
  \centering
  \includegraphics[scale=1]{figure1.pdf}
  \caption{\textbf{Cultural horizontal transmission with assortment.} Transmission occurs between interacting partners with probability $\alpha$ (left) or between two random peers with probability $1-\alpha$, where $\alpha$ is the \emph{social association} parameter.
  }
  \label{fig:horizontal}
\end{figure}

%%% Figure: coexistence map - without oblique transmission
\begin{figure}[h]
\centering
\includegraphics[scale = 0.55]{coexistence_without_oblique.pdf}
\caption{\textbf{Stable and unstable coexistence between cooperation and defection without oblique transmission.}
  The curves show the frequency of the cooperative phenotype $A$ among parents in the next generation, $\tilde{p}'$, vs.\ that in the current generation $\tilde{p}$ (\autoref{eq:nextgen_parents_vertical_only}).
  The dashed black line is $\tilde{p}'=\tilde{p}$.
  The curves and the dashed line intersect at the polymorphic equilibrium $\tilde{p}^*$ (black circle).
  When the curves are above the dashed line, $\tilde{p}'>\tilde{p}$, and $\tilde{p}$ increases.
  When the curves are below the dashed line, $\tilde{p}'<\tilde{p}$, and $\tilde{p}$ decreases.
  The orange curve, for which the polymorphic equilibrium is stable, is given by $T_A = 0.4$, $T_B = 0.9$, $b = 12$, $c=0.35$, and $\alpha = 0.45$, which give $\gamma_2<c<\gamma_1$ (\autoref{eq:cost_boundaries})
  The blue curve, for which the equilibrium is unstable, is given by $T_A = 0.5$, $T_B = 0.1$, $b = 1.3$, $c=0.904$, and $\alpha = 0.4$, which give $\gamma_1<c<\gamma_2$.
  In both cases there is no oblique transmission, $v=1$; see \autoref{fig:coexistence_with_oblique} for $v<1$.
  }
\label{fig:coexistence_without_oblique}
\end{figure}

@@

\begin{figure}[h]
  \centering       
    \includegraphics[width=0.45\textwidth]{Result2_c_zero_alpha.eps}
	~
    \includegraphics[width=0.45\textwidth]{Result2_c_non_zero_alpha.eps}
    ~
    \includegraphics[width=0.45\textwidth]{Result2_alpha.eps}

    \caption{\textbf{Evolution of cooperation under vertical and horizontal cultural transmission.} 
The figure shows the global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable coexistence (blue), and stable coexistence of cooperation and defection (yellow).
In all cases the horizontal bias ($T_A-T_B$) is on the x-axis.
(\textbf{a-b})~The cost of cooperation $c$ is on the y-axis; the cost boundaries $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_boundaries}) are the solid and dashed lines, respectively. 
(\textbf{c})~social association $\alpha$ is on the y-axis; the social association boundaries $a_1$ and $a_2$ (\autoref{eq:boundries_assortative_meeting}) are the solid and dashed lines, respectively.
Here, $b=1.3$, $T_A=0.4$. (\textbf{a})~$\alpha = 0$.(\textbf{b})~$\alpha = 0.7$. (\textbf{c})~$c = 0.35$.
  	}
    \label{fig:result2}
\end{figure}

@@

\begin{figure}[h]
  \centering       
    \includegraphics[width=0.45\textwidth]{Result3_TA_TB.eps}~
    \includegraphics[width=0.45\textwidth]{Result3_TB_TA.eps}
    \includegraphics[width=0.45\textwidth]{Result3_alpha_Vs_v_TA_TB.eps}~
    \includegraphics[width=0.45\textwidth]{Result3_alpha_Vs_v_TB_TA.eps}
    \caption{\textbf{Evolution of cooperation under vertical, oblique, and horizontal cultural transmission.} 
    The figure shows the global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable coexistence (blue), and stable coexistence of cooperation and defection (yellow).
	In all cases the vertical transmission rate $v$ is on the x-axis.
	(\textbf{a-b}) The cost of cooperation $c$ is on the y-axis and the cost boundaries $\hat\gamma_1$ and $\hat\gamma_2$ (\autoref{eq:cost_boundaries_v}) are represented by the solid and dashed lines, respectively. 
    (\textbf{c-d}) The social association $\alpha$ is on the y-axis and the social association boundaries $\hat{a}_1$ and $\hat{a}_2$ (\autoref{eq:boundries_assortative_meeting_general_case}) are represented by the solid and dashed lines, respectively. 
    Horizontal transmission is biased in (\textbf{a,c}) for cooperation, $T_A>T_B$, and in (\textbf{b,d}) for defection, $T_A<T_B$.    
    Here, $T_A = 0.5$, and
    (\textbf{a}) $b=1.2$, , $T_B = 0.4$, $\alpha = 0.4$;
    (\textbf{b}) $b=2$, $T_B = 0.7$, $\alpha = 0.7$;
    (\textbf{c}) $b=1.2$, $T_B = 0.4$, $c=0.5$;
    (\textbf{d}) $b=2$, $T_B = 0.7$, $c=0.5$.
    }
    \label{fig:result3}
\end{figure}
